A pipeline to analyse time-course gene expression data

### 2019-05-14 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\hypertarget{pre-processing-and-normalization}{%
\subsubsection{Pre-processing and normalization}\label{pre-processing-and-normalization}}

\textbf{EAP: Should we make this part of the pipeline, or just start with normalization? It's very microarray heavy\ldots{} We could have a vignette in \texttt{moanin} documenting the creation of a normalized version of the data.}

The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
with accession number GSE63786. The dataset has a total of 209 samples, with
45018 probes (1 sample was removed by the authors prior to uploading the
dataset on GEO).

Note that in the following, we assume that the user has access to a data
folder \texttt{data/shoemaker2015/rawdata}, containing the microarray data.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data_dir =}\StringTok{ "data/shoemaker2015/rawdata"}
\end{Highlighting}
\end{Shaded}

Given the source and a \texttt{Targets.txt} structured as follows, microarray can be
loaded thanks to the limma library.

\begin{verbatim}
FileName
GSM1557140_AR0313_043raw.txt.gz
GSM1557141_AR0313_044raw.txt.gz
GSM1557142_AR0313_045raw.txt.gz
GSM1557143_AR0313_055raw.txt.gz
GSM1557144_AR0313_056raw.txt.gz
GSM1557145_AR0313_057raw.txt.gz
GSM1557146_AR0313_058raw.txt.gz
GSM1557147_AR0313_059raw.txt.gz
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{targets =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{readTargets}\NormalTok{()}
\NormalTok{RG =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{read.maimages}\NormalTok{(}
\NormalTok{  targets, }\DataTypeTok{source=}\StringTok{"agilent"}\NormalTok{,}
  \DataTypeTok{path=}\NormalTok{data_dir,}
  \DataTypeTok{other.columns=}\StringTok{"gIsWellAboveBG"}\NormalTok{,}
  \DataTypeTok{green.only=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The normalization strategy we apply here relies on \texttt{limma} and is as follows:

\begin{itemize}
\tightlist
\item
  First, normalize probe intensity using background information using the
  ``norm-exponential'' strategy.
\item
  Then normalize across micro-arrrays using quantile normalization.
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Perform background correction using norm-exponential strategy and normalize}
\CommentTok{# across samples using quantile normalization.}
\NormalTok{RG =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{backgroundCorrect}\NormalTok{(RG, }\DataTypeTok{method=}\StringTok{"normexp"}\NormalTok{)}
\NormalTok{RG.q =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{normalizeBetweenArrays}\NormalTok{(RG, }\DataTypeTok{method=}\StringTok{"quantile"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Plotting the smoothed empirical densities of the probe values before
normalization shows considerable variations betweens arrays. Once the data has
been normalized across array using quantile normalization, the smoothed
empirical densities of each sample perfectly overlap.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Plot the densities of the probes before and after normalization.}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{))}
\NormalTok{limma}\OperatorTok{::}\KeywordTok{plotDensities}\NormalTok{(RG, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{legend=}\OtherTok{FALSE}\NormalTok{,}
             \DataTypeTok{main=}\StringTok{"Before normalization"}\NormalTok{, }\DataTypeTok{cex=}\FloatTok{0.8}\NormalTok{)}
\NormalTok{limma}\OperatorTok{::}\KeywordTok{plotDensities}\NormalTok{(RG.q, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{legend=}\OtherTok{FALSE}\NormalTok{,}
             \DataTypeTok{main=}\StringTok{"After normalization"}\NormalTok{, }\DataTypeTok{cex=}\FloatTok{0.8}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/plotting_densities-1} \end{center}

We then clean up the names of each gene and each sample to match the original
data. Finally, we further filter out probes based on the following criteria:
(1) removing the 1486 control probes; (2) removing
4823 low-expressed genes.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sample_names =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\KeywordTok{strsplit}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(RG.q), }\StringTok{"_"}\NormalTok{), .subset2,  }\DecValTok{1}\NormalTok{)}
\KeywordTok{colnames}\NormalTok{(RG.q) =}\StringTok{ }\NormalTok{sample_names}
\KeywordTok{row.names}\NormalTok{(RG.q) =}\StringTok{ }\KeywordTok{make.names}\NormalTok{(RG}\OperatorTok{$}\NormalTok{genes}\OperatorTok{$}\NormalTok{SystematicName, }\DataTypeTok{unique=}\OtherTok{TRUE}\NormalTok{)}
\NormalTok{idx_to_keep =}\StringTok{ }\KeywordTok{which}\NormalTok{(}
  \OperatorTok{!}\NormalTok{(RG}\OperatorTok{$}\NormalTok{genes}\OperatorTok{$}\NormalTok{Systematic }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"DarkCorner"}\NormalTok{, }\StringTok{"GE_BrightCorner"}\NormalTok{,}
                   \StringTok{"NegativeControl"}\NormalTok{)))}

\NormalTok{data =}\StringTok{ }\NormalTok{RG.q[idx_to_keep,]}

\CommentTok{# Filter out control probes.}
\NormalTok{control_probes =}\StringTok{ }\NormalTok{data}\OperatorTok{$}\NormalTok{genes[, }\StringTok{"ControlType"}\NormalTok{] }\OperatorTok{==}\StringTok{ }\DecValTok{1}
\NormalTok{is_expressed =}\StringTok{ }\KeywordTok{rowSums}\NormalTok{(data}\OperatorTok{$}\NormalTok{other}\OperatorTok{$}\StringTok{"gIsWellAboveBG"} \OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{>=}\StringTok{ }\DecValTok{4}

\NormalTok{data =}\StringTok{ }\NormalTok{data[}\OperatorTok{!}\NormalTok{control_probes }\OperatorTok{&}\StringTok{ }\NormalTok{is_expressed, ]}\OperatorTok{$}\NormalTok{E}
\end{Highlighting}
\end{Shaded}

After filtering, we are left with 39544 probes of interest in
209 samples.

Now that we have normalized and filtered the gene expressions, let's load the
metadata. The package \texttt{moanin} contains the normalized data and meta of
\citep{shoemaker:ultrasensitive}.

@@

Plotting the smoothed empirical densities of probe values for each sample and
estimating the variations across samples before and after normalization is a
first quality control step. \textbf{EAP: Cut or move this sentence, so that this entire section is applicable to microarray or mRNA-Seq}

Typically, two other quality control and
exploratory analysis steps are also performed before and after normalization:
(1) low dimensionality embedding of the samples; (2) correlation plots between
each samples. In both cases, we expect a strong biological signal, while
replicate samples should be strongly clustered or correlated with one another.

@@

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-12-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}

@@

@@

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-2} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-32-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-33-1} \end{center}

>>>>

First let's load the data. The package \texttt{moanin} contains the normalized
data and meta of \citep{shoemaker:ultrasensitive}.

@@

Typically, two quality control and exploratory analysis steps are also
performed before and after normalization: (1) low dimensionality embedding of
the samples; (2) correlation plots between each samples. In both cases, we
expect a strong biological signal, while replicate samples should be strongly
clustered or correlated with one another.

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-5-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-8-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-16-1} \end{center}

@@

@@

@@

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-2} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}

@@

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
